Visual narratives of the Covid-19 pandemic

### 2022-01-24 JDSSV_manuscript_knitr.Rnw
Remove orphaned figures

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

@@

Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage. Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region.

@@

And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
\caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
\end{subfigure}\hfill
\begin{subfigure}[t]{.45\textwidth}
\includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
\caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
\end{subfigure}
\end{figure}

>>>>

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:

\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

@@

Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage. Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region. 
Clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.

@@

@@

% \begin{figure}
% \centering
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{covid-tracking-hex-cartogram}
% \caption{Hexagonal Cartogram of Covid-19 cases in the United States by state, per million people. This somewhat mitigates the area bias issue, but is still subject to other perceptual biases. Source: The COVID Tracking Project at The Atlantic \url{https://covidtracking.com/data/charts/us-daily-positive-cases}.}\label{fig:hex-cartogram}
% \end{subfigure}\hfill
% \begin{subfigure}[t]{.45\textwidth}
% \includegraphics[width=\textwidth]{nyt-covid-scaled-choro}
% \caption{Choropleth Map of covid cases per country, scaled by population. Source: New York Times \url{https://www.nytimes.com/interactive/2021/world/covid-cases.html}.}\label{fig:scaled-choro}
% \end{subfigure}
% \caption{Modifications of choropleth maps which attempt to address the area bias of choropleths in different ways.}
% \end{figure}

### 2022-01-24 JDSSV_manuscript_knitr.tex
Remove orphaned figures

The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message is a challenge. It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.

>>>>

The pandemic and graphics used to show the ebb and flow of COVID cases highlight the challenges of data visualization in the modern era: even with numerous guidelines and standards for graphics, it is not straightforward to communicate about complex data, and creating charts which fully represent the intended message or messages is a challenge. It is essential that we continue to study how graphics are perceived and interpreted, supporting the guidelines and standards which exist with user studies that ground these rules in science.
